Tidal streams in triaxial systems

### 2015-03-04 ms.tex
werq it

In this work, we analyze whether the final configuration-space morphology of simple tidal stream models is correlated with measures of chaos local to the orbit of the progenitor. The effect of chaos in the halos of galaxies has largely been neglected because the chaotic timescales estimated were thought to be many times the age of the universe, but recent work has shown that even weak chaos may be relevant for tidal stream morphology. Here we choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics, and compare these results to measures of the mixing time of finite-volume ensembles of orbits (analogs to tidal debris). Finally, we connect the ensemble mixing rate to observational properties, thus demonstrating that local orbital measures of chaos can be used to predict the [...].

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/orbit-tree.pdf}
\caption{} \label{fig:orbit-tree}
\end{center}
\end{figure*}

@@

For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori -- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}).

@@

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
where $\delta_i=\|\delta \bs{w}(t_i)\|$ and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}

@@

NAFF has been applied to many problems in Solar System and Galactic dynamics, especially in the study of of orbits in triaxial systems. For more details about the method, see \cite{papaphilippou96, laskar, etc.}. We have implemented and tested our own implementation of NAFF in Python. This code is publicly available through the Gary package.\footnote{link to repo} Figure~\ref{fig:logfreqs} shows a test of the implementation in which we reproduce the frequency map at a fixed energy of an axisymmetric, logarithmic potential \cite[][pg. 260, Figure~3.45]{binneytremaine}. Plotted are the (Cartesian) frequency ratios recovered for a grid of iso-energy, box orbits integrated in the potential
\begin{equation}
	\Phi(x,y,z) = \frac{1}{2}\ln\left(x^2 + (y/0.9)^2 + (z/0.7)^2 + 0.1\right).
\end{equation}
In such a map, stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities.

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/log-freqmap.png}
\caption{This is really just a placeholder that shows what I want to show here...} \label{fig:logfreqs}
\end{center}
\end{figure*}

\subsection{Comparison of methods}

We choose two nearby short-axis tube orbits in the potential described in Section~\ref{sec:potential} in order to demonstrate the two methods described above. The orbits were chosen from the grids of initial conditions described in later sections because they are very close in phase-space but display very different behavior. In particular, Orbit A is regular and Orbit B is stochastic. Figure~\ref{fig:two-orbits} shows the two orbits in projection -- Orbit A is shown in the left column and Orbit B the right. Two consecutive integration periods for each orbit are shown as the two rows -- note the change in appearance of Orbit B (the chaotic orbit, right column). The initial conditions for the two orbits in physical units and Cartesian coordinates are:
\begin{align}
	\bs{x}_{0}^{(A)} &= (27.85, 0, 23.35)~{\rm kpc}\\
	\bs{v}_{0}^{(A)} &= (0, 115.37, 0)~{\rm km~s}^{-1}\\
	\bs{x}_{0}^{(B)} &= (27.85, 0, 23.10)~{\rm kpc}\\
	\bs{v}_{0}^{(B)} &= (0, 116.43, 0)~{\rm km~s}^{-1}.
\end{align}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/two-orbits.pdf}
\caption{ } \label{fig:two-orbits}
\end{center}
\end{figure*}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/lyap-orbits.pdf}
\caption{ } \label{fig:two-methods}
\end{center}
\end{figure*}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/freq-diff-orbits.pdf}
\caption{ } \label{fig:fdiff-orbits}
\end{center}
\end{figure*}

Figure~\ref{fig:two-methods} shows $l_N$ computed for the two orbits. Over-plotted on both panels are lines (dashed) that show $t^{-0.9}$ behavior. The MLE estimate continues towards zero for the regular orbit (top) but saturates for the chaotic orbit (bottom). For mildly chaotic or ``sticky'' orbits, it can take integration periods of many thousands of orbits to determine a reliable estimate of the Lyapunov exponent. In comparison, owing to the precision of NAFF, detecting small frequency diffusion rates can be done with only $\sim$100 orbital periods. Figure~\ref{fig:fdiff-orbits} shows the three fundamental frequencies computed for 12 consecutive windows of orbit (each section is $\sim$100 orbital periods in length) for Orbit A (left) and Orbit B (right). Dashed horizontal line is the estimate of the uncertainty in the numerical computation of the frequencies from Orbit A. The value of each frequency in the first window is subtracted from each subsequent window. The frequencies only show minor deviations for Orbit A (left panel) whereas there is significant diffusion in Orbit B (right panel).

@@

The Lyapunov time and frequency diffusion rate are local measures of chaos but ...

We therefore take a simplified approach and consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event of a low-mass progenitor. For a given set of orbital initial conditions -- the parent orbit -- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all of these test particles for some evolution time, $T_{\rm ev}$. For a given mass scale, the ensemble particle position and velocity offsets are assumed to be Normal-distributed away from the parent orbit with spreads set by the tidal radius in position and velocity scale in velocity \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and is drawn from
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, and $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above scales.

As the particle ensembles evolve in time, we follow \citet[][hereafter MV96]{merritt96} and quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniform over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because that is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small ($\approx$1\%) spreads in energy, but we still expect that such an ensemble around a parent chaotic orbit will reach a fully mixed state. The final distribution, in addition to filling the energy hypersurface, will also have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble will never reach the fully mixed state as defined above because it will remain confined into a subspace on the energy surface. In practice, only strongly chaotic systems will approach $\rho_{E_0}(\bs{x})$; the rate will depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate), and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is 
\begin{align}
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, \rho_0 \, \delta(H - E_0)\\
	&= \int \dd^3v \, \rho_0 \, \delta\left(|\bs{v}|^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= 4\pi \int \dd v \, v^2 \, \rho_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.

Second, we compute the Kullback-Leibler divergence \citep[KL divergence;][]{kullback51} by estimating the configuration space density $\rho(\bs{x},t_j)$ using kernel density estimation (KDE)\footnote{We use an implementation from the Python package scikit-learn \citep{scikitlearn}.} with the ensemble of particle positions at time $t_j$, then use Monte Carlo integration to evaluate the KL divergence rather than evaluate the integral over a three dimensional grid. The KL divergence, $D_{\rm KL}$, is a measure of the relative Shannon entropy between two normalized density functions, $p(x)$ and $q(x)$,
\begin{equation}
	D_{\rm KL} = \int \dd x \, p(x) \ln \frac{p(x)}{q(x)}
\end{equation}
We would like to compute 
\begin{align}
	D_{\rm KL}(t_j) &= \int \dd^3 \bs{x} \, \rho(\bs{x},t_j) \ln \frac{\rho(\bs{x},t_j)}{\rho_{E_0}(\bs{x})}
\end{align}
however, we only have an estimate of the density $\rho_{\rm KDE}(\bs{x},t_j) \approx \rho(\bs{x},t_j)$. Using samples from this KDE estimated density, $\bs{x}_i$, the KL divergence is
\begin{align}
	D_{\rm KL}(t_j) &\approx \frac{1}{N} \sum_i^N \ln \frac{\rho_{\rm KDE}(\bs{x}_i,t_j)}{\rho_{E_0}(\bs{x}_i)}\\
	\bs{x}_i &\sim \rho_{\rm KDE}(\bs{x},t_j)\label{eq:sampled_from}
\end{align}
where Equation~\ref{eq:sampled_from} indicates that the $\bs{x}_i$ are samples from the KDE density. For any computation of the density estimate, we use 10-fold cross-validation to find the optimal kernel bandwidth, $\sigma_{\rm KDE}$, by maximizing the likelihood of withheld particles.

Figure~\ref{fig:DKL} shows $D_{\rm KL}$ as a function of time for a regular (black) and chaotic (blue) orbit. The chaotic orbit ensemble mixes much faster than the regular orbit ensemble.

@@

We start by studying the properties of isoenergy orbits in the potential described above. We generate grids of initial condition along the $xz$ ($y=0$) and $yz$ ($x=0$) planes\footnote{Recall that $x$ is the major, $y$ the intermediate, and $z$ the minor axis.} all with the same energy, $E=XXX$. For the $xz$ orbits, we fix $v_x = v_z = 0$, and compute $v_y$ from the energy. Similarly, for the $yz$ orbits, we fix $v_y = v_z = 0$, and compute $v_x$ from the energy. We integrate the orbits as described in Section~\ref{sec:integration} and compute the fundamental frequencies of each orbit with NAFF. Figure~\ref{fig:freqmap} shows frequency maps for the $xz$ (left) and $yz$ (right) grids. For short-axis tubes, the frequencies plotted are the radial frequency, $\Omega_r$, the azimuthal frequency around the $z$ axis, $\Omega_\phi$, and the oscillation along the $z$ axis, $\Omega_z$. For long-axis tubes, the frequencies plotted are the radial frequency, $\Omega_r$, the azimuthal frequency around the $x$ axis, $\Omega_\phi$, and the oscillation along the $x$ axis, $\Omega_x$. Box orbits are not shown. Already evident in these panels is that many of the orbits are regular: the transformed, even-spaced sampling of the initial conditions is preserved in the frequency transformation, implying that the frequencies for many orbits do not change appreciably. Also apparent is resonant trapping,  which appear as over-dense, linear features in these frequency maps.

>>>>

In this work, we analyze whether the final configuration-space morphology of simple tidal stream models is correlated with measures of chaos local to the orbit of the progenitor. The effect of chaos in the halos of galaxies has largely been neglected because the chaotic timescales estimated were thought to be many times the age of the universe, but recent work has shown that even weak chaos may be relevant for tidal stream morphology. Here we choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics, and compare these results to measures of the mixing time of finite-volume ensembles of orbits (analogs to tidal debris). Finally, we connect the ensemble mixing rate to observational properties of streams, thus demonstrating that local orbital measures of chaos are predictors of tidal stream morphology. This provides a strong case for developing a method to use the mere existence of the many thin streams found in the halo of our own Galaxy to rule out classes or shapes of possible potentials that fit the Milky Way.

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
\end{center}
\end{figure*}

@@

For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}).

@@

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i \label{eq:ftmle}
\end{equation}
where $\delta_i=\|\delta \bs{w}(t_i)\|$ and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}

@@

NAFF has been applied to many problems in Solar System and Galactic dynamics, especially in the study of of orbits in triaxial systems. For more details about the method, see \cite{papaphilippou96, laskar, etc.}. We have implemented and tested our own implementation of NAFF in Python. This code is publicly available through the Gary package.\footnote{link to repo} Figure~\ref{fig:logfreqs} shows a test of the implementation in which we reproduce the frequency map at a fixed energy of an axisymmetric, logarithmic potential \cite[][pg. 260, Figure~3.45]{binneytremaine}. Plotted are the (Cartesian) frequency ratios recovered for a grid of iso-energy, box orbits integrated in the potential
\begin{equation}
	\Phi(x,y,z) = \frac{1}{2}\ln\left(x^2 + (y/0.9)^2 + (z/0.7)^2 + 0.1\right). \label{eq:logpotential}
\end{equation}
In such a map, stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/log-freqmap.png}
\caption{Ratio of cartesian frequencies for a grid of iso-energy box orbits integrated in the potential of Equation~\ref{eq:logpotential}. Stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.} \label{fig:logfreqs}
\end{center}
\end{figure*}

NAFF recovers the fundamental frequencies for an orbit faster (with a fewer number of terms) when the coordinates used are ``close'' to the angle variables \cite[PL96][]{papaphilippou96}. PL96 show that a good choice of coordinates for tube orbits are the Poincar\'e symplectic polar coordinates, a set of canonical coordinates similar to cylindrical coordinates. When computing the frequencies for tube orbits, we first rotated the orbits until the circulation is about the $z$-axis, transform to Poincar\'e polar coordinates, then use NAFF to measure the fundamental frequencies. We could equivalently use the Cartesian time series, but the convergence of terms is slower. We have tested that our implementation of NAFF returns the same frequencies in either case for a set of tube orbits. For box orbits, the motion is close to separable in each Cartesian component and we therefore use Cartesian coordinates for estimating frequencies for these orbits.

\subsection{Comparison of methods}\label{sec:comparison}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/two-orbits.pdf}
\caption{ Two orbits integrated in the triaxial potential described in Section~\ref{sec:potential}. Left panels (black) show two consecutive sections of orbit, where the integration time for each section (rows) is set to 100 times the maximum orbital period. Right panels (red) shows the same for the chaotic orbit. The appearance of the orbit in configuration space changes appreciably over timescales of tens of orbital periods. } \label{fig:two-orbits}
\end{center}
\end{figure*}

We choose two nearby short-axis tube orbits in the potential described in Section~\ref{sec:potential} in order to demonstrate the two methods described above. The orbits were chosen from a grid of initial conditions in the $x$-$z$ plane (described in later sections) because they are very close in phase-space but display very different behavior. In particular, Orbit A is regular or at least weakly chaotic and Orbit B is strongly chaotic. Figure~\ref{fig:two-orbits} shows the two orbits in projection --- Orbit A is shown in the left column and Orbit B the right. Two consecutive integration periods for each orbit are shown in the two rows; note the change in appearance of Orbit B (right column). The initial conditions for the two orbits in physical units and Cartesian coordinates are:
\begin{align}
	\bs{x}_{0}^{(A)} &= (27.85, 0, 23.35)~{\rm kpc}\\
	\bs{v}_{0}^{(A)} &= (0, 115.37, 0)~{\rm km~s}^{-1}\\
	\bs{x}_{0}^{(B)} &= (27.85, 0, 23.10)~{\rm kpc}\\
	\bs{v}_{0}^{(B)} &= (0, 116.43, 0)~{\rm km~s}^{-1}.
\end{align}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.62\textwidth]{figures/lyap-orbits.pdf}
\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over 1000 orbital periods (the maximum orbital period) for Orbit A (left) and Orbit B (right). For Orbit A, the FTMLE asymptotes towards zero with a characteristic power-law slope. For Orbit B, the FTMLE begins to saturate to a constant; the Lyapunov exponent estimated for this orbit is $\lambda_{\rm max} \approx 0.07 \, {\rm Gyr}^{-1}$ corresponding to a Lyapunov time of $t_\lambda \approx 14 \, {Gyr}$. Over-plotted on both panels are lines (dashed, blue) that show $t^{-0.9}$ behavior.} \label{fig:lyap-orbits}
\end{center}
\end{figure*}
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.62\textwidth]{figures/freq-diff-orbits.pdf}
\caption{ Difference in frequencies computed for successive sections of 100 orbital periods from the frequencies computed in the initial section for Orbit A (left) and Orbit B (right). The three lines in each panel are for the three fundamental frequencies. Dashed (red) horizontal line shows the estimate of the uncertainty for the frequencies computed for Orbit A. After many hundreds of orbital periods, the frequency deviation is not appreciable indicating again that the orbit is regular or only very weakly chaotic. The frequency deviations computed for Orbit B, however, show a clear increasing trend, indicating that the frequencies are diffusing with time. } \label{fig:fdiff-orbits} % The frequency diffusion rate defined in Equation~\ref{eq:fdrate} measures the initial slope of this trend.
\end{center}
\end{figure*}

Figure~\ref{fig:lyap-orbits} shows $l_N$ computed for the two orbits described above. The MLE estimate continues towards zero for the regular orbit (left) but saturates for the chaotic orbit (right). For mildly chaotic or ``sticky'' orbits, it can take integration periods of many thousands of orbits to determine a reliable estimate of the Lyapunov exponent. In comparison, owing to the precision of NAFF, detecting small frequency diffusion rates can be done with only $\sim$100 orbital periods. Figure~\ref{fig:fdiff-orbits} shows the three fundamental frequencies computed for 12 consecutive windows of orbit (each section is $\sim$100 orbital periods in length) for Orbit A (left) and Orbit B (right). Dashed horizontal line is the estimate of the uncertainty in the numerical computation of the frequencies from Orbit A. The value of each frequency in the first window is subtracted from each subsequent window. The frequencies only show minor deviations for Orbit A (left panel) whereas there is significant diffusion in Orbit B (right panel).

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit. That is, they are measures of how infinitesimal deviations will diverge, on average, from some parent orbit of interest. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. This raises an important consideration: do finite deviations diverge faster than predicted from local measures of chaos? Does the Lyapunov time or frequency rate predict the timescale over which a finite phase-space volume stays ``coherent''?

It is computationally intractable to run full N-body simulations of stellar system disruption for large grids of orbital initial conditions and thus we take a simplified approach for analyzing these questions. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the parent orbit --- we find the position of the nearest pericenter, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all of these test particles for some evolution time, $T_{\rm ev}$. The physical scale of the ensemble is set by the tidal radius in position and velocity scale in velocity and are therefore set by the mass scale of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. For this work, we take $m=2\times10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass to estimate the above scales.

As the particle ensembles evolve in time, we follow a method similar to that of \citet[][hereafter MV96]{merritt96} and quantify the degree of mixing of the ensemble by computing the ``distance'' between the configuration space density of the ensemble at a given time, $\rho(\bs{x},t)$, with that of a distribution that is uniform over the energy hypersurface, $\rho_{E_0}(\bs{x})$, where $E_0$ is the energy of the parent orbit and thus hypersurface of interest. Our method differs in a few ways from that of MV96. First, MV96 use isoenergy ensembles of particles for following the density because this is the only way to guarantee that, if chaotic, nearly all orbits in the ensemble will lie on chaotic orbits. We instead use ensembles with small, finite spreads in energy, but we still expect that most ensemble particle orbits around a parent chaotic orbit will reach a fully mixed state. The final distribution, in addition to filling the energy hypersurface, will also have some thickness around this surface. Around a typical regular orbit, each individual orbit will fill its torus ergodically --- a surface with lower dimension than the energy hypersurface --- and thus, a small enough ensemble will never reach the fully mixed state as defined above because it will remain confined into a subspace of lower dimensionality. In practice, only strongly chaotic systems will get close to $\rho_{E_0}(\bs{x})$; we expect the rate to depend on how stochastic the parent orbit is (e.g., as measured by the Lyapunov time or frequency diffusion rate) and also on how ``trapped'' the chaotic region is (a small chaotic region wedged between stable tori will take longer to appear chaotic, even if internally the frequency diffusion rate is large). However the difference in configuration space density between an ensemble undergoing chaotic mixing versus an ensemble around a regular orbit may be detectable over a much shorter timescale. The fully-mixed, normalized, configuration-space density that fills a given energy hypersurface, $E_0$, is 
\begin{align}
	\rho_{E_0}(\bs{x}) &= \int \dd^3v \, \rho_0 \, \delta(H - E_0)\\
	&= \int \dd^3v \, \rho_0 \, \delta\left(|\bs{v}|^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= 4\pi \int \dd v \, v^2 \, \rho_0 \, \delta\left(v^2/2 + \Phi(\bs{x}) - E_0\right)\\
	&= A \sqrt{E_0 - \Phi(\bs{x})}\label{eq:mixeddensity}
\end{align}
where $A$ is a constant determined such that $\int \dd^3\bs{x} \, \rho_{E_0}(\bs{x}) = 1$, and $\delta(\cdot)$ is the Dirac delta function.

For each ensemble, we compute the Kullback-Leibler divergence \citep[KL divergence;][]{kullback51} between the configuration-space density and the fully-mixed density (Equation~\ref{eq:mixeddensity}) by estimating the configuration space density $\rho(\bs{x},t_j)$ using kernel density estimation (KDE)\footnote{We use an implementation from the Python package scikit-learn \citep{scikitlearn}.} with the ensemble of particle positions at a given time $t_j$, We use Monte Carlo integration to evaluate the KL divergence rather than evaluate the integral over a three dimensional grid. The KL divergence, $D_{\rm KL}$, is a measure of the relative (Shannon) entropy between two normalized density functions, $p(x)$ and $q(x)$,
\begin{equation}
	D_{\rm KL} = \int \dd x \, p(x) \ln \frac{p(x)}{q(x)}
\end{equation}
We would like to compute 
\begin{align}
	D_{\rm KL}(t_j) &= \int \dd^3 \bs{x} \, \rho(\bs{x},t_j) \ln \frac{\rho(\bs{x},t_j)}{\rho_{E_0}(\bs{x})}
\end{align}
however, we only have an estimate of the density $\rho_{\rm KDE}(\bs{x},t_j) \approx \rho(\bs{x},t_j)$. Using samples from this KDE estimated density, $\bs{x}_i$, the KL divergence is
\begin{align}
	D_{\rm KL}(t_j) &\approx \frac{1}{N} \sum_i^N \ln \frac{\rho_{\rm KDE}(\bs{x}_i,t_j)}{\rho_{E_0}(\bs{x}_i)}\\
	\bs{x}_i &\sim \rho_{\rm KDE}(\bs{x},t_j)\label{eq:sampled_from}
\end{align}
where Equation~\ref{eq:sampled_from} indicates that the $\bs{x}_i$ are samples from the KDE density. For any computation of the density estimate, we use 10-fold cross-validation to find the optimal kernel bandwidth, $\sigma_{\rm KDE}$, by maximizing the likelihood of withheld particles. [TODO: previous sentence?]

Figure~\ref{fig:DKL} shows $D_{\rm KL}$ as a function of time for the two orbits of Section~\ref{sec:comparison}. The chaotic orbit (Orbit B) ensemble mixes much faster than the regular orbit (Orbit A) ensemble, but note the timescale here is still over many tens of orbital periods. The KLD is related to the log-density of the debris so that the final difference in KLD shown in Figure~\ref{fig:DKL} corresponds to a factor of 10 difference in density. TODO: From the decay of the KLD, we measure a mixing rate, defined as [...] TODO: what is the mixing rate? etc...

@@

We start by studying the properties of isoenergy orbits in the potential described above. We generate isoenergy grids of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy $E=XXX$. We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits --- the other major class of orbits in triaxial potentials --- which tend to plunge deep into the inner regions of the potential. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the fast disruption expected for stellar systems on radially plunging orbits. The major classes of stable tube orbits circulate about either the major or minor axis. The most prominent orbits in this grid are the short-axis and outer long-axis tubes, but this grid generates all of the major orbit classes: short-axis tubes, long-axis inner tubes, long-axis outer tubes, stochastic intermediate-axis, and box orbits. [TODO: Stochastic tube orbits not well studied...because not very important here...]

We integrate all orbits for $\approx$200 orbital periods (Section~\ref{sec:integration}) and use NAFF to compute the fundamental frequencies for the two consecutive sections of 100 orbital periods. Figure~\ref{fig:freqmap} (left) shows a map of the fundamental frequency ratios for all tube orbits from the input grid (we plot the frequencies estimated from the first 100 orbital periods). Already evident in this panel is that many of the orbits are regular: the transformed, even-spaced sampling of the initial conditions is preserved in the frequency transformation, implying that the frequencies for many orbits do not change appreciably. Also apparent is resonant trapping, which appear as over-dense, linear features. For approximately XX\% of the orbits, NAFF fails to recover the frequencies (similar difficulties were reported in \cite{valluri98} and \cite{papaphilippou96} for recovering certain tube orbit frequencies).
